---
title: "Practical Machine Learning Course Project"
author: "Davis Balaba"
date: "Sunday, May 24, 2015"
output: html_document
---

Step 1 : Load  and inspect data

```{r}
library(caret)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(randomForest)
library(rattle)

train <- read.csv("C:/Users/davis.balaba/Desktop/Coursera/Practical Machine Learning/pml-training.csv",na.strings=c("NA","#DIV/0!", ""), header=TRUE)

test <- read.csv("C:/Users/davis.balaba/Desktop/Coursera/Practical Machine Learning/pml-testing.csv",na.strings=c("NA","#DIV/0!", ""), header=TRUE)

#colnames(train)


```
We observe that columns 1 to 7 should be eliminated as they are not pertinent to the prediction process.

Step 2: Clean data

```{r}
train <- train[,8:length(colnames(train))]
dim(train)
test <- test[,8:length(colnames(test))]

# Delete columns with all missing values
training2<-train[,colSums(is.na(train)) == 0]
testing2 <-test[,colSums(is.na(test)) == 0]
dim(training2)
dim(testing2)

```

Step 3: Split training sample 60/40 to create in-sample test set
```{r, echo=FALSE}
set.seed(90)
inTrain = createDataPartition(training2$classe, p = .60)[[1]]
train_tr = training2[ inTrain,]
train_test = training2[-inTrain,]

dim(train_tr)
dim(train_test)
```

Step 4a: Modeling using decision tree
We examine the output variable and observe that it is a 5-level categorical variable. So first we attempt a decision tree model.
```{r, echo=FALSE}

summary(train$classe)
set.seed(90)
#Create model
model1 <- rpart(classe ~ ., data=train_tr, method="class")
# Predict
prediction1 <- predict(model1, train_test, type = "class")
#Evaluate model
# Test results on our subTesting data set:
confusionMatrix(prediction1, train_test$classe)

```
This approach yields ~73% accuracy.  

Step 4b: Modeling using random forest
```{r}
set.seed(90)
#Create model
model2 <- randomForest(classe ~ ., data=train_tr, method="class")
# Predict
prediction2 <- predict(model2, train_test, type = "class")
#Evaluate model
# Test results on our subTesting data set:
confusionMatrix(prediction2, train_test$classe)

```

The random forest yields about 99% accuracy and will therefore be used for the final predictions. 

Step 5: Final Prediction
```{r}

prediction3 <- predict(model2, testing2, type="class")
prediction3

```
